---
title: "Schedule"

output:
  distill::distill_article:
    self_contained: true
    toc: true
    toc_float: true
---

```{r, include=FALSE}
require(here)
require(kableExtra)

knitr::opts_chunk$set(echo = FALSE,
                      message = FALSE,
                      warning = FALSE)

tab <- read.csv(here('data/schedule.csv'))
```

```{r}

tbl <- function(tab,d){
  
  loc <- which(tab$date==d)
  
  tab2 <- tab[loc,]
  
  if(!is.na(tab2$notes)){
    
    tab2$notes_       <- paste0(cell_spec(strsplit(tab2$notes,';')[[1]],
                                          'html',
                                          link=strsplit(tab2$notes_link,';')[[1]]),
                                collapse = '<br>')
  } else {
    tab2$notes_ <- ''
  }
  
  
  if(!is.na(tab2$kaggle)){
    
    tab2$kaggle_      <- paste0(cell_spec(strsplit(tab2$kaggle,';')[[1]],
                                          'html',
                                          link=strsplit(tab2$kaggle_link,';')[[1]]),
                                collapse = '<br>')
  } else {
    
    tab2$kaggle_ <- ''
    
  }
  
  
  if(!is.na(tab2$slides)){
      
    tab2$slides_      <- paste0(cell_spec(strsplit(tab2$slides,';')[[1]],
                                          'html',
                                          link=strsplit(tab2$slides_link,';')[[1]]),
                                collapse = '<br>')
  } else {
    tab2$slides_ <- ''
  }
  
  
  
  if(!is.na(tab2$reading)){
     
    tab2$reading_     <- paste0(cell_spec(strsplit(tab2$reading,';')[[1]],
                                            'html',
                                            link=strsplit(tab2$reading_link,';')[[1]]),
                                  collapse = '<br>')
  } else{
    tab2$reading_ <- ''
  }
  
  if(!is.na(tab2$assignments)){
    
    tab2$assignments_ <- paste0(cell_spec(strsplit(tab2$assignments,';')[[1]],
                                            'html',
                                            link=strsplit(tab2$assignments_link,';')[[1]]),
                                  collapse = '<br>')
  } else{
    tab2$assignments_ <- ''
  }
  
  
  tab2 <- tab2[,c('date','notes_','kaggle_','slides_','reading_','assignments_')]
  
  rownames(tab2) <- NULL
  
  tab2 %>%
    kbl(format = 'html',
        escape = FALSE,
        col.names = c('Date','Notes','Kaggle Notebooks','Slides','Optional Supplemental Readings','Assignments')) %>%
    kable_material(c("striped", "hover"),full_width = T) %>%
    row_spec(0, background = "#F6FBF4") %>%
    row_spec(0:1, extra_css = "border-bottom: solid;") %>%
    row_spec(0:1, extra_css = "border-top: solid;") %>%
    row_spec(0:1, extra_css = "border-left: solid;") %>%
    row_spec(0:1, extra_css = "border-right: solid;")
  
}
```

### Week 1: Introduction

```{r}

tbl(tab,d='3-Oct')

```

### Week 2: Data Preprocessing

```{r}

tbl(tab,d='10-Oct')

```

### Week 3 & Week 4: Introduction to Linear Regression, Bias/Variance Tradeoff, and Cross-validation

```{r}

tbl(tab,d='17-Oct & 24-Oct')

```

### Week 5: Regularized Linear Regression
```{r}

tbl(tab,d='31-Oct')

```

### Week 6 & 7: (Regularized) Logistic Regression
```{r}

tbl(tab,d='7-Nov & 14-Nov')

```

### Week 8: Introduction to K-Nearest Neighbors and Decision Tree Algorithms
```{r}

tbl(tab,d='21-Nov')

```

### Week 9: Introduction to Bagged Trees, Random Forests, and Gradient Boosting Trees
```{r}

tbl(tab,d='28-Nov')

```

